Experience Report: Haskell in Computational Biology

### 2012-02-15 mrfy_experience_report.tex
added diagram, MRFy architecture explanation. Some revisions to introduction.

\usepackage{amsmath}
\usepackage{array}
\usepackage{listings}
\lstset{
    tabsize = 2,
    basicstyle = \ttfamily,
    language = Haskell
    }

@@

Computational biologists write software to solve problems on large genomic or 
proteomic data sets. Traditionally, these software programs fall into two 
categories: those with high-performance computing needs are typically 
implemented in C++, while the remainder are typically implemented in Perl, 
Python, Ruby, or R. Recently, we have developed a program in Haskell which 
implements a novel algorithm for solving the remote homology detection problem. 
While designing the algorithm, we wondered: would the algorithm be easily 
expressed in Haskell? Would we be able to achieve acceptable runtime 
performance? How would our experience compare with our past experiences 
developing and maintaining computational biology software in C++, Ruby, and 
Python? What would be our greatest challenges? In this paper, we briefly 
explain the remote homology detection problem and our algorithm, and then 
attempt to answer the above questions.

As a brief refresher in the standard dogma of genetics, recall that genes 
(strands of deoxyribonucleic acid (DNA), which are sequences of the nucleotides 
adenosine, cytosine, guanine, and thymine, represented by the letters A, C, G, 
and T) are transcribed into ribonucleic acid (RNA). Some RNA is then translated 
into the 20 naturally-occuring amino acids, which form peptide chains -- 
proteins -- which fold into complex structures in the cell. These proteins are 
cellular machinery, performing the varied functions an organism needs to 
survive. Since the dawn of the human genome project in 1990, the cost of 
sequencing an entire genome has dropped exponentially. Recently, the doubling 
time for genomic data has been one year, in contrast to the eighteen month 
doubling time of Moore's law. However, determining what newly sequenced genes 
actually do -- when they become proteins, what structures those proteins fold 
into, and what functions they perform -- has not kept pace. Determining the 
atomic coordinates of a newly discovered protein may require as much as two 
graduate-student-years in a lab!

@@

However, when there are no known proteins of similar sequence, only 
evolutionarily distant proteins that may share similar structure, this problem 
is called remote homology detection. Hidden Markov model approaches begin to 
fail as sequence similarity falls off. Recently, Markov random field 
approaches, which capture non-local interactions in the protein structure, have 
shown to perform well. However, these approaches face the challenge of 
increased computational complexity; the SMURF program exhibits exponential time 
complexity, while SMURFLite (work by one of the authors) bounds the exponent by 
simplifying the Markov random field's dependency graph. We have implemented a 
stochastic search approach to Markov random fields, called MRFy.

The algorithm underlying MRFy grew out of an understanding of the limitations 
of the SMURF (Structural Motifs Using Random Fields) algorithm, which used 
multidimensional dynamic programming to compute the optimal parse of a query 
sequence onto a Markov random field.

TODO: (add explanation of MRFy architecture and algorithm)

>>>>

\usepackage{amsmath}
\usepackage{array}
\usepackage{listings}
\usepackage{graphicx}
\lstset{
    tabsize = 2,
    basicstyle = \ttfamily,
    language = Haskell
    }

@@

Computational biologists write software that answers questions about large data 
sets involving sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). This software comes in two varieties. The first variety 
is \textit{high performance software}, where performance is a top priority; 
this kind of software is usually written in a programming language like C or 
C++. The second variety is software where performance is not a top priority; 
this kind of software written in higher-level languages that emphasize 
convenience and programmer productivity, such as Perl, Python, Ruby, or R.

In this Experience Report, we show that the Haskell programming language is an 
excellent alternative for computational biologists: a Haskell program can 
perform comparably to a C++ program while being as flexible and easy to 
understand as a program written in an interpreted language. Our experience is 
based on an implementation of a novel algorithm for finding remote homologs to 
newly discovered proteins. We call this program MRFy.

As a brief refresher in the standard dogma of genetics, recall that genes 
(strands of deoxyribonucleic acid (DNA), which are sequences of the nucleotides 
adenosine, cytosine, guanine, and thymine, represented by the letters A, C, G, 
and T) are transcribed into ribonucleic acid (RNA). Some RNA is then translated 
into the 20 naturally-occuring amino acids, which form peptide chains -- 
proteins -- which fold into complex structures in the cell. These proteins are 
cellular machinery, performing the varied functions an organism needs to 
survive.

However, determining what newly sequenced genes actually do -- when they become 
proteins, what structures those proteins fold into, and what functions they 
perform -- has not kept pace. Determining the atomic coordinates of a newly 
discovered protein may require as much as two graduate-student-years in a lab! 
Fortunately, determining the full structure of a protein is not always 
necessary. Often, we can identify well-known proteins of similar sequence or 
structure to a new protein, and thus make educated guesses as to the structure 
and function of the new protein.

@@

However, when there are no known proteins of similar sequence, only 
evolutionarily distant proteins that may share similar structure, this problem 
is called remote homology detection. Hidden Markov model approaches begin to 
fail as sequence similarity falls off. Recently, Markov random field 
approaches, which capture non-local interactions in the protein structure, have 
been shown to perform well. However, these approaches face the challenge of 
increased computational complexity; the SMURF program exhibits exponential time 
complexity, while SMURFLite (work by one of the authors) bounds the exponent by 
simplifying the Markov random field's dependency graph. MRFy is a 
stochastic search approach to Markov random fields.

SMURF, SMURFLite, and MRFy all attempt to compute the optimal parse, or 
alignment, of a protein sequence onto the directed graph representing a Markov 
random field (see Figure \ref{mrf}). The algorithm underlying MRFy grew out of 
an understanding of the limitations of the SMURF (Structural Motifs Using 
Random Fields) algorithm, which used multidimensional dynamic programming to 
compute the optimal parse of a query sequence onto a Markov random field. 
Rather than attempt to compute exactly the optimal parse, MRFy treats the 
non-local interactions (which correspond to hydrogen-bonded ``beta strands'' in 
the protein) as ``beads'' which can be slid along a string. In this analogy, 
the string is the sequence of amino acids of the query protein. A given 
placement of these beads corresponds to an assignment of particular amino acid 
residues being hydrogen bonded to other particular amino acid residues, and 
thus MRFy computes a probability based on observed frequencies of residues 
hydrogen bonding in beta strands~\cite{betawrappro}. In between these beta 
strands, the Markov random field is in fact a traditional hidden Markov model, 
whose optimal solution can be easily computed via dynamic programming (known as 
the Viterbi algorithm). MRFy performs a stochastic search to place these beta 
strands, based on an objective function that combines the probability computed 
by the Viterbi algorithm with that computed from the paired, hydrogen-bonded 
residues in the beta strands.

\begin{figure}
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}}
\caption{The directed graph representing the states of a MRFy Markov random field finite state
automaton}\label{mrf}
\end{figure}
